Recurrence spike spectra

### 2022-05-18 Manuscript/inter_spike_manuscript.tex
Logistic example updated

We consider the Logistic map $x_{n+1}=r\cdot x_n \left( 1-x_n \right)$ for changing control parameter $r$. We vary $r$ from $r=3.4$ to $r=4$ in steps of $0.001$. For 
each setting of $r$ 
\begin{itemize}[noitemsep]
\item[(1)] a time series of length $N=201$ is computed with a random initial condition $u_0 \in [0,\ 1]$, neglecting the first $1,000$ samples as transients,
\item[(2)] $100$ iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000} are computed,
\item[(3)] the time series and its iAAFT surrogates are embedded in a 2-dimensional state space using a time delay of unity,
\item[(4)] from the 2-dimensional trajectories RPs, Eq.~\eqref{eq_rp_definition}, are computed under a threshold $\varepsilon=0.05$,
\item[(5)] $\tau$-RR, Eq.~\eqref{eq_tau_rr}, is computed from the RP of the signal and from the RPs of the surrogates,
\item[(6)] inter spike spectra are obtained from $\tau$-RR of the signal and from the $\tau$-RRs of the surrogates, see Section \ref{sec_tau_rr_method}, and finally,
\item[(7)] from the distribution of the surrogate inter spike spectra the $95^\text{th}$ percentile is computed. The peaks of the inter spike spectrum of the signal which exceed 
this percentile are counted. 
\end{itemize}
In this example, the according null hypothesis is that the data stems from a process which yields the same auto-correlation, hence the same Fourier powerspectrum, and the same 
amplitude distribution. We consider the number of significant peaks in the inter spike spectrum with
respect to the control parameter in order to distinguish the corresponding dynamics (Fig.~\ref{fig_tau_rr_logistic}C).
A correlation with the positive Lyapunov exponent (Fig.~\ref{fig_tau_rr_logistic}A) is discernible 
($\rho_{\text{Pearson}}=0.88$). Moreover, this analysis can tackle period-doubling, since it ``measures'' the dominant cycles via the inter spike spectrum.

A less computationally intensive approach is to compute surrogates for the $\tau$-RR analytically, rather than computing a RP and its $\tau$-RR for each iAAFT surrogate of the 
time series. This translates into a null hypothesis 
that the $\tau$-RR and its corresponding inter spike spectrum stems from a RP of a random signal. In this case the probability of finding a black point in the RP can be obtained 
from a binomial distribution with probability parameter $p$ set to the recurrence rate of the RP of the signal. This way $100$ surrogate $\tau$-RRs are computed in step (5). 
The results are similar to the ones obtained from the iAAFT surrogates and nearly as good (Fig.~\ref{fig_tau_rr_logistic}B). The first period doubling at 
$r \approx 3.458$ cannot be detected, while the correlation to the positive Lyapunov exponent is only slightly worse ($\rho_{\text{Pearson}}=0.84$).

The described procedure does work well for map data, because most often the $\tau$-RR for those kind of data reveals a ``spiky enough'' nature. 
On the contrary, highly sampled (flow-) data often yield not as 
spiky $\tau$-RR's and, thus, the number of significant peaks in the inter spike spectrum may not be sensitive enough to detect period-doubling bifurcations. Nevertheless the 
according inter spike spectra is still revealing important information (Fig.~\ref{fig_tau_rr_example_roessler}) and practitioners can design appropriate quantifying statistics based 
on these spectra, which suit the 
research task.

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_tau_rr_logistic}
 \caption{\textbf{A} Lyapunov exponent of the Logistic map as a function of the control parameter $r$. 
 \textbf{B} Number of significant peaks ($\alpha=0.05$) in the inter spike spectrum of the $\tau$-RR, random surrogates. 
 \textbf{C} Same as \textbf{B}, but for iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000}.
}
\label{fig_tau_rr_logistic}
\end{figure}

>>>>

We consider the Logistic map $x_{n+1}=r\cdot x_n \left( 1-x_n \right)$ for changing control parameter $r$. We vary $r$ from $r=3.4$ to $r=4$ in steps of $0.001$. For 
each setting of $r$ 
\begin{itemize}[noitemsep]
\item[(1)] a time series of length $N=201$ is computed with a random initial condition $u_0 \in [0,\ 1]$, neglecting the first $1,000$ samples as transients,
\item[(2)] $100$ iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000} are computed,
\item[(3)] the time series and its iAAFT surrogates are embedded in a 2-dimensional state space using a time delay of unity,
\item[(4)] from the 2-dimensional trajectories RPs, Eq.~\eqref{eq_rp_definition}, are computed under a threshold $\varepsilon=0.05$,
\item[(5)] $\tau$-RR, Eq.~\eqref{eq_tau_rr}, is computed from the RP of the signal and from the RPs of the surrogates,
\item[(6)] inter spike spectra are obtained from $\tau$-RR of the signal and from the $\tau$-RRs of the surrogates, see Section \ref{sec_tau_rr_method}, and finally,
\item[(7)] from the distribution of the surrogate inter spike spectra the $95^\text{th}$ percentile is computed. The peaks of the inter spike spectrum of the signal which exceed 
this percentile are counted. 
\end{itemize}
In this example, the according null hypothesis is that the data stems from a process which yields the same auto-correlation, hence the same Fourier powerspectrum, and the same 
amplitude distribution. We consider the number of significant peaks in the inter spike spectrum with
respect to the control parameter in order to distinguish the corresponding dynamics (Fig.~\ref{fig_tau_rr_logistic}C).
A correlation with the positive Lyapunov exponent (Fig.~\ref{fig_tau_rr_logistic}A) is discernible 
($\rho_{\text{Pearson}}=0.72$). Moreover, this analysis can tackle period-doubling, since it ``measures'' the dominant cycles via the inter spike spectrum.

A less computationally intensive approach is to compute surrogates for the $\tau$-RR analytically, rather than computing a RP and its $\tau$-RR for each iAAFT surrogate of the 
time series. This translates into a null hypothesis 
that the $\tau$-RR and its corresponding inter spike spectrum stems from a RP of a random signal. In this case the probability of finding a black point in the RP can be obtained 
from a binomial distribution with probability parameter $p$ set to the recurrence rate of the RP of the signal. This way $100$ surrogate $\tau$-RRs are computed in step (5). 
The results are even slightly better compared to the ones obtained from the iAAFT surrogates (Fig.~\ref{fig_tau_rr_logistic}B, $\rho_{\text{Pearson}}=0.81$). The first period doubling at 
$r \approx 3.458$ cannot be detected by any of the surrogates.

The described procedure does work well for map data, because most often the $\tau$-RR for those kind of data reveals a ``spiky enough'' nature. 
On the contrary, highly sampled (flow-) data often yield not as 
spiky $\tau$-RR's and, thus, the number of significant peaks in the inter spike spectrum may not be sensitive enough to detect period-doubling bifurcations. Moreover the sensitivity of the 
inter spike spectrum to detect the regime shifts also depends on the critical regularization threshold. Nevertheless the 
according inter spike spectra is still revealing important information (Fig.~\ref{fig_tau_rr_example_roessler}) and practitioners can design appropriate quantifying statistics based 
on these spectra, which suit the 
research task.

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_tau_rr_logistic}
 \caption{\textbf{A} Lyapunov exponent of the Logistic map as a function of the control parameter $r$. 
 \textbf{B} Number of significant peaks ($\alpha=0.05$) in the inter spike spectrum of the $\tau$-RR and its Pearson correlation coefficient to the Lyapunov exponent shown in \textbf{A} 
 (white noise surrogates). 
 \textbf{C} Same as \textbf{B}, but for iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000}. For obtaining the inter spike spectra we used 
 a regularization threshold corresponding to $\rho=0.95$ accordance of $\tau$-RR's and re-composed signals.  
}
\label{fig_tau_rr_logistic}
\end{figure}
